Negative Types

### 2014-05-08 negatives.tex
not sure

\newcommand{\geqz}{\geq\zerog}
\newcommand{\leqz}{\leq\zerog}
\newcommand{\zerog}{\mathbb{0}}
\newcommand{\oneg}{\mathbb{1}}
\newcommand{\twog}{\mathbb{2}}
\newcommand{\threeg}{\mathbb{3}}
\newcommand{\fourg}{\mathbb{4}}
\newcommand{\fiveg}{\mathbb{5}}

@@

@@

\textbf{Comparing numbers.} There are three outcomes to games represented by
Conway numbers $\mathbb{n}$:
\begin{itemize}
\item $\mathbb{n} > \zerog$ which indicates that the \emph{left} player can
  enforce a win, no matter who starts;
\item $\mathbb{n} < \zerog$ which indicates that the \emph{right} player can
  enforce a win, no matter who starts;
\item $\mathbb{n} = \zerog$ which indicates that the \emph{second}
  player can enforce a win, no matter who.
\end{itemize}
For example, we have $\threeg > \zerog$ because the right player would have
no options (i.e., would lose) on the first or second turn depending on who
starts. Similarly, we have $-\threeg < \zerog$ because the left player would
have no options on on the first or second turn depending on who starts. For
$\zerog$, it is evident that the first player to take a turn must lose, i.e.,
that the second player wins. More interestingly, consider the game
$\textsf{CN}~[-\oneg]~[\oneg]$ which we claim is $= \zerog$. Indeed, if the
left player starts, the only option is to move to the game $-\oneg$ which the
right player wins. If however the right player moves first, the only option
is to move to the game $\oneg$ which the left player wins. In both cases, the
second player wins. To ensure that these three outcomes are the only possible
outcomes, Conway numbers have an additional restriction: for every left
option $x$ and right option $y$, we require that $x < y$. If that restriction
is lifted we could construct games such that $\textsf{CN}~[\zerog]~[\zerog]$
for which none of the relations $> \zerog$, $< \zerog$, or $= \zerog$
holds. Instead for this game, the first player always wins, which means that
the left player cannot enforce a win (in case the right player starts) and
vice-versa.

@@

\noindent\textbf{Addition and subtraction.} Addition of two games intuitively
gives each player the choice of selecting an option from either
game. Subtraction is simply the addition of the dual of a game. Formally:
\begin{code}
(:+:) :: ConwayNumber -> ConwayNumber -> ConwayNumber
x@(CN xls xrs) :+: y@(CN yls yrs) = 
  CN 
    ([ xl :+: y | xl <- xls ] `union`
     [ x :+: yl | yl <- yls ])
    ([ xr :+: y | xr <- xrs ] `union`
     [ x :+: yr | yr <- yrs ])

@@

\noindent\textbf{Multiplication.} The most complicated operation on Conway
numbers is multiplication. It is defined as follows:
\begin{code}
(:*:) :: ConwayNumber -> ConwayNumber -> ConwayNumber
x@(CN xls xrs) :*: y@(CN yls yrs) = 
  CN 
    ([ (xl :*: y) :+: (x :*: yl) :-: (xl :*: yl)
     | xl <- xls, yl <- yls] `union`
     [ (xr :*: y) :+: (x :*: yr) :-: (xr :*: yr)
     | xr <- xrs, yr <- yrs])
    ([ (xl :*: y) :+: (x :*: yr) :-: (xl :*: yr)
     | xl <- xls, yr <- yrs] `union`
     [ (xr :*: y) :+: (x :*: yl) :-: (xr :*: yl)
     | xr <- xrs, yl <- yls])
\end{code}

@@

>>>>

\newcommand{\geqz}{\geq\zerog}
\newcommand{\leqz}{\leq\zerog}
\newcommand{\zerog}{\mathbb{0}}
\newcommand{\oneg}{\mathbb{1}}
\newcommand{\twog}{\mathbb{2}}
\newcommand{\threeg}{\mathbb{3}}
\newcommand{\fourg}{\mathbb{4}}
\newcommand{\fiveg}{\mathbb{5}}
\newcommand{\bck}{\triangleleft}
\newcommand{\fwd}{\triangleright}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{$\Pi$: Operational and Denotational Semantics}

Present $\Pi$ with an interpreter specifying the operational semantics. For a
model or denotational semantics, the minimum requirement is that every type
maps to some kind of space; a value $v : t$ maps to an element of that space;
and a combinator $c : t_1 \leftrightarrow t_2$ maps to a function between the
spaces denoting $t_1$ and $t_2$. The simplest model one can think of is to
map a type of size $n$ to a finite set of $n$ elements; a value $v : t$ maps
to an element in the set denoting $t$; a combinator maps to a permutation on
that set. One could also build other models, e.g. vector spaces but that
model is fine. Perhaps show correspondence between operational and
denotational semantics.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{$\Pi^{-}$: Operational Semantics}

Extend $\Pi$ with negative types and $\eta$ and $\epsilon$. Present the
operational semantics that uses two interpreters forwards and backwards.

For the model or denotational semantics, we need to map every type to some
space. If type $t$ maps to space $S$, we need to have an operation that maps
that space to another space that acts like ``negative'' $S$ so that we can
satisfy for example that $S-S=0$. Using the cardinality of the set like we
did for $\Pi$ does not directly work (footnote on papers that describe sets
with negative cardinality). Another idea is to generalize spaces to have two
components (a producer and a consumer; a left player and a right player) and
to describe the space using the relative ``strength'' of the two
components. (This is the idea of the Int or G construction but that doesn't
directly work if we have two monoidal structures.) A natural and beautiful
theory is that of Conway games which model two player games: left and
right. In that setting we can talk about games in which the left player has
an $n$-move advantage (summarized by the number $n$) and games in which the
right player has an $n$-move advantage (summarized by the number $-n$ which
denotes that the left player has an $n$-move disadvantage). It is therefore
possible to map a type of size $n$ where $n$ is an integer to the class
Conway games where the left player has an $n$-move (dis-)advantage; the
negative of a space $S$ is the space in which all the games are flipped with
left and right players exchanging positions; a value $v : t$ maps to a Conway
number in the set denoting $t$; all combinators preserve the size of the type
and hence will be the identity. Note that the space corresponds to type $0$
is not empty in this denotation: it consists of all the games in which the
left player has a $0$-move advantage over the the right player and
symmetrically, i.e., the set of games which are always lost by the first
player to move.

Given that our language is reversible, a backward evaluator is relatively
straightforward to implement: using the backward evaluator to calculate
$c(v)$ is equivalent to $c^{\dagger}(v)$ in the forward evaluator.

\begin{figure*}
\[\begin{array}{rclr}
[\mathit{iso}, v, C, s]_{\bck} &\longmapsto& 
  \langle \mathit{iso}, v', C, s'\rangle_{\bck} 
  \quad\textrm{where}~\mathit{iso}^{\dagger}~s~v \longmapsto (v', s') \\
\langle c_1, v, \mathit{Fst}~C~c_2, s \rangle_{\bck} &\longmapsto& 
  \langle c_1 \fatsemi c_2, v, C, s \rangle_{\bck} \\
\langle c_2, v, \mathit{Snd}~c_1~C, s \rangle_{\bck} &\longmapsto& 
  [c_1, v, \mathit{Fst}~C~c_2, s]_{\bck} \\
[ c_1 \fatsemi c_2, v, C, s ]_{\bck} &\longmapsto& 
  [c_2, v, \mathit{Snd}~c_1~C, s]_{\bck} \\
\langle c_1, v, \mathit{LeftP}~C~c_2, s \rangle_{\bck} &\longmapsto& 
  \langle c_1 \oplus c_2, \mathit{left}~v, C \rangle_{\bck} \\
[c_1 \oplus c_2, v', C, s ]_{\bck} &\longmapsto& 
  [ c_1, v, \mathit{LeftP}~C~c_2, s[v' \approx left v] ]_{\bck} \\
\langle c_2, v, \mathit{RightP}~c_1~C, s\rangle_{\bck} &\longmapsto& 
  \langle c_1 \oplus c_2, \mathit{right}~v, C \rangle_{\bck} \\
[c_1 \oplus c_2, v', C,s ]_{\bck} &\longmapsto& 
  [ c_2, v, \mathit{RightP}~c_1~C, s[v' \approx right v] ]_{\bck} \\
\langle c_1, v_1, \mathit{LeftT}~C~c_2~v_2, s \rangle_{\bck} &\longmapsto& 
  \langle c_1 \otimes c_2, (v_1, v_2), C, s \rangle_{\bck} \\
\langle c_2, v_2, \mathit{RightT}~c_1~v_1~C, s \rangle_{\bck} &\longmapsto& 
  [ c_1, v_1, \mathit{LeftT}~C~c_2~v_2, s ]_{\bck} \\
[ c_1 \otimes c_2, v, C, s ]_{\bck} &\longmapsto& 
  [ c_2, v_2, \mathit{RightT}~c_1~v_1~C, S' ]_{\bck} 
  \quad\textrm{where}~ s' = s[v \approx (v_1, v_2)] \\
\\
\\
\langle \epsilon, v, C, s \rangle_{\fwd} &\longmapsto& 
  \langle \epsilon, \mathit{left}~(-v'), C, s[v \approx right v'] \rangle_{\bck} \\
\langle \epsilon, v, C, s \rangle_{\fwd} &\longmapsto& 
  \langle \epsilon, \mathit{right}~v', C, s[v \approx left (-v')] \rangle_{\bck} \\
\\
\\
\langle \eta, v, C, s \rangle_{\bck} &\longmapsto& 
  \langle \eta, \mathit{left}~(-v'), C, s[v \approx right v'] \rangle_{\fwd} \\
\langle \eta, v, C, s \rangle_{\bck} &\longmapsto& 
  \langle \eta, \mathit{right}~v', C, s[v \approx left (-v')] \rangle_{\fwd}
\end{array}\]
\end{figure*}

We add the following rules to the reductions above. 
\begin{enumerate}
\item The rules for $\epsilon$ essentially transfer control from the forward
  evaluator (whose states are tagged by $\triangleright$) to the backward
  evaluator (whose states are tagged by $\triangleleft$). In other words,
  after an $\epsilon$ the direction of the world is reversed. The pattern
  matching done by the unification ensures that a value on the right wire is
  tagged to be negative and transferred to the left wire, and vice versa.
  Note that there is no evaluation rule for $\eta$ in the forward
  evaluator. This corresponds to the fact that there is no value of type 0
  and hence the forward evaluator can never execute an $\eta$.

\item The rules for $eta$ are added to the backward evaluator. A program
  executing backwards starts executing forwards after the execution of the
  $\eta$. Dual to the previous case, there is no rule for $\epsilon$ in the
  backward evaluator since the output type of $\epsilon$ is 0.
\end{enumerate}

@@

\smallskip\noindent\textbf{Comparing numbers.} There are three outcomes to
games represented by Conway numbers $\mathbb{n}$:
\begin{itemize}
\item $\mathbb{n} > \zerog$ which indicates that the \emph{left} player can
  enforce a win, no matter who starts;
\item $\mathbb{n} < \zerog$ which indicates that the \emph{right} player can
  enforce a win, no matter who starts;
\item $\mathbb{n} = \zerog$ which indicates that the \emph{second}
  player can enforce a win, no matter who.
\end{itemize}
For example, we have $\threeg > \zerog$ because the right player would have
no options (i.e., would lose) on the first or second turn depending on who
starts. Similarly, we have $-\threeg < \zerog$ because the left player would
have no options on on the first or second turn depending on who starts. For
$\zerog$, it is evident that the first player to take a turn must lose, i.e.,
that the second player wins. More interestingly, consider the game
$\textsf{CN}~[-\oneg]~[\oneg]$ which we claim is $= \zerog$. Indeed, if the
left player starts, the only option is to move to the game $-\oneg$ which the
right player wins. If however the right player moves first, the only option
is to move to the game $\oneg$ which the left player wins. In both cases, the
second player wins. To ensure that these three outcomes are the only possible
outcomes, Conway numbers have an additional restriction: for every left
option $x$ and right option $y$, we require that $x < y$. If that restriction
is lifted we could construct games such that $\textsf{CN}~[\zerog]~[\zerog]$
for which none of the relations $> \zerog$, $< \zerog$, or $= \zerog$
holds. Instead for this game, the first player always wins, which means that
the left player cannot enforce a win (in case the right player starts) and
vice-versa.

@@

\smallskip\noindent\textbf{Addition and subtraction.} Addition of two games
intuitively gives each player the choice of selecting an option from either
game. Subtraction is simply the addition of the dual of a game. Formally:
\begin{code}
(:+:) :: ConwayNumber -> ConwayNumber -> ConwayNumber
x@(CN xls xrs) :+: y@(CN yls yrs) = 
  CN 
    ([ xl :+: y | xl <- xls ] `union`
     [ x :+: yl | yl <- yls ])
    ([ xr :+: y | xr <- xrs ] `union`
     [ x :+: yr | yr <- yrs ])

@@

\smallskip\noindent\textbf{Multiplication.} The most complicated operation on
Conway numbers is multiplication. It is defined as follows:
\begin{code}
(:*:) :: ConwayNumber -> ConwayNumber -> ConwayNumber
x@(CN xls xrs) :*: y@(CN yls yrs) = 
  CN 
    ([ (xl :*: y) :+: (x :*: yl) :-: (xl :*: yl)
     | xl <- xls, yl <- yls] `union`
     [ (xr :*: y) :+: (x :*: yr) :-: (xr :*: yr)
     | xr <- xrs, yr <- yrs])
    ([ (xl :*: y) :+: (x :*: yr) :-: (xl :*: yr)
     | xl <- xls, yr <- yrs] `union`
     [ (xr :*: y) :+: (x :*: yl) :-: (xr :*: yl)
     | xr <- xrs, yl <- yls])
\end{code}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Correspondence}

Show correspondence between operational and denotational semantics for
$\Pi^{-}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
